Complementary roles of Theta and Alpha in Multi-Item Associations

### 2019-12-16 manuscript.tex
chaning to apa style
revisions with replication

%Dear all,
%We ar writing to invite you to submit an article to an upcoming special issue of the journal Hippocampus with the tentative title: Computational models of the hippocampus and related structures.  The theme includes computational and theoretical models as well as theory-driven data analysis of the hippocampus and related cortical and sub-cortical structures. 
%
%Articles for the special issue will be due June 15, 2019 with a planned publication date in early 2020.  All articles will be published on-line when accepted, but we will need to be firm about the deadline in order to avoid delays in publishing the full issue with volume and page numbers.
% 
%
%The contributions to the special issue can be in all forms of manuscripts accepted by Hippocampus (Commentaries, Reviews, Articles, Rapid Communications).
% 
%
%As is common with these types of special issues, we ask that any contributing author commit to reviewing two articles for the special issue.
% 
%
%Please let me know by Dec. 15 if you plan to contribute to this special issue.  Please also send me a tentative title and potential author list (which can change later if necessary).
%
%Best regards,
%Michael Hasselmo
%and Ehren Newman

@@

\usepackage{subcaption}\usepackage[style=numeric,backend=bibtex8]{biblatex}
\addbibresource{references.bib}
\usepackage{longtable}
\usepackage{csvsimple}				% for reading cvs files into tables

@@

%\section*{Acknowledgements}\label{sec:acknowledge}

@@

It is generally accepted that long-term memory encoding and retrieval are not isolated functions of any one brain area \cite{Gabrieli98,PrinceDaselaarCabeza05}.  However, there are many challenges in moving beyond the traditional studies focusing on individual regions  (e.g., concentrating on the prominent role of the hippocampus), to obtain insight into the nature of network-level dynamics.  An increasing body of work support the idea that oscillatory dynamics in multiple brain regions are associated with different aspects of  memory encoding and retrieval \cite{NyhusCurran10,HanslmayrStaudiglFellner12,Klimesch12}, %update refs here!!!
These studies have generally found a consistent relationship between several frequency bands and successful encoding or retrieval of experiences; namely that oscillatory power within the alpha (8 to 12Hz) and beta (13 to 30Hz) frequency bands generally decrease with the successful execution of these memory processes, while theta (3 to 8Hz) and gamma (30 to 100+Hz) increase in the regions directly involved in the memory operations. Why these relationships exist, and what they tell us about the larger functional organization of memory systems in the brain are still open questions.

One model suggests that thalamic circuits play a functional role in selectively gating communication between cortical regions by synchronizing them in various frequency bands \cite{ShermanGuillery06,SaalmannKastner11,KetzJensenOReilly15}.  These gating mechanisms, implemented  by oscillatory synchronization, are central to system level brain function and ultimately affect cognitive behavior such as memory encoding and retrieval.  Further, the particular oscillatory dynamics engaged by this selection process may provide indicators of which system-scale neural circuits are coordinating together to facilitate that particular form of memory processing.

A particular thalmo-cortical circuit of interest for memory processes involves the hippocampus and its associated subcortical network, including the anterior thalamus, and medial septum which are essential in the maintenance of theta oscillations (4 to 8 Hz) \cite{Buzsaki02}.  Theta oscillations have been shown to be critically involved in successful encoding and retrieval of episodic like memories in both humans and animals \cite{NyhusCurran10,JonesWilson05,FuentemillaBarnesDuzelEtAl14}, and are believed to be primarily driven by the pacemaker cells within the medial septum.  The anterior thalamus, however,  provides the connectivity for the hippocampus to potentially modulate communication with the medial and orbital PFC through theta synchronization \cite{VanderWerfJollesWitterEtAl03,AggletonDumontWarburton11}.  Thus, the theta circuit is important for the core functioning of the hippocampus during encoding and retrieval.

%There is MEG evidence that theta synchronization with the hippocampus can facilitate Recollective memory processes in humans \cite{FuentemillaBarnesDuzelEtAl14}. There is some indirect MEG and BOLD evidence that the thalamus plays a role in this synchronization process \cite{FuentemillaBarnesDuzelEtAl14,PergolaRanftMathiasEtAl13}, which is bolstered by the known connectivity between the hippocampus and PFC through the anterior thalamus from rodents \cite{AggletonBrown99,AggletonDumontWarburton11}.  The thalamic gating model connects these ideas suggesting Recollective memory works through the hippocampus via theta synchronization, however to date there is no strong evidence supporting this claim in humans.

A complimentary thalamic circuit involves the pulvinar providing synchronization within the alpha (8 to 12 Hz) band between parietal and visual cortices with the medial, orbital and lateral PFC \cite{BarbasHenionDermon91,KievitKuypers77,LopesdaSilva91}.  This circuit is important for bottom-up stimulus-driven encoding processes in memory \cite{JutrasFriesBuffalo13}, but it can also inhibit contradictory or interfering information during internally guided retrieval \cite{ParkLeeKangEtAl14}.  Thus, its overall contribution to memory can be mixed, and depends on the nature of the task and what kinds of memory processing is required.

%The pulvinar has a well established role in coordinating alpha synchrony in animal models \cite{SaalmannKastner11,SaalmannKastner09,LopesdaSilva91,LopesdaSilvaVosMooibroekEtAl80}, and there is recent evidence in human models showing alpha oscillations provide an active function of suppressing cortex that would otherwise interfere with the memory retrieval \cite{WaldhauserJohanssonHanslmayr12,HanslmayrStaudiglAslanEtAl10,KhaderRosler11}.  There is also human fMRI evidence showing PFC modulation of visual cortex through the pulvinar \cite{SotoHumphreysRotshtein07,RotshteinSotoGrecucciEtAl11}.

Indeed, a recent spiking model of these oscillations called the Sync/deSync Model highlights the role of a synchronized hippocampus that accompanies a desynchronized neocortex during encoding and retrieval.  This model in particular focuses on the role of decreased alpha power related to successful encoding of new information \cite{ParishHanslmayrBowman18, HanslmayrStaresinaBowman16}.  Similarly, recent empirical studies of human intracranial recordings find evidence to support the complimentary roles of hippocampal synchrony and neocortical desynchrony in successful encoding and retrieval \cite{GriffithsMichelmannRouxEtAl18, StaresinaMichelmannBonnefondEtAl16}.  In general these results find a decrease in neocortical alpha power and an increase in hippocampal gamma power that coincide with successful encoding and retrieval of declarative memories.

This accumulated evidence suggests a relationship between theta and alpha frequency bands and long-term memory systems as constructed by the Complimentary Learning Systems theory \cite{McClellandMcNaughtonOReilly95,OReillyBhattacharyyaHowardEtAl14}.  Namely, that theta power positively tracks successful engagement (encoding/retrieval) between hippocampus and neocortex, while alpha power negatively tracks with this process as active inhibition of cortex is released in response to the exchange of information between hippocampus and neocortex.  This relationship is further supported by the connectivity of the thalamus with cortex such that a specific model for how information is routed into and out of these systems can be tied to empirical measures of neural oscillations.  This work seeks to test this relationship through a series of computational simulations of hippocampal-thalamo-cortical interaction, and an accompanying EEG study in humans to compare against simulation results.

@@

\subsection{Dependent Events Paradigm}
%
%Hippocampal dependent memory is generally thought to be associative in nature, where multiple cortical representations are bound together in a distinct hippocampal representation during encoding.  Later, an incomplete set of those cortical representations can be used as a cue to reactive the full set during retrieval \cite{McClellandMcNaughtonOReillyEtAl92}.  This multi-element type of memory retrieval is also referred to behaviorally as `Recollection' where multiple details of a given experience can be retrieved, and hippocampal activity increases with subjective confidence and the amount of information being retrieved \cite{RuggVilbergMattsonEtAl12}.  
%
%In contrast to Recollection, memories that exhibit `Familiarity' have no measurable details that can be recalled, however individual elements can be successfully recognized when presented with a mix of lures.  This Familiarity process is thought to rely more heavily on the medial temporal lobe (MTL) cortex surrounding the hippocampus, including entorhrinal, parahippocampal, and perirhinal cortices.  These cortical areas, among others, are proposed to be a separate memory system from the hippocampus, serving behavior when recall of all associated details is not necessary or available \cite{EichenbaumYonelinasRanganath07,AggletonBrown99}.  This theory of separate memory systems has been contentious for several decades with evidence both in support and in contrast to its core ideas  \cite{EichenbaumYonelinasRanganath07,WixtedSquire10}; namely that Recollection can occur in addition to Familiarity and that the hippocampus is not specific to Recollection but can be used in Familiarity based processing.  What is clear is that MTL cortex learns and maintains representations required for hippocampal retrieval, and those MTL regions along with the hippocampus can be used to re-instantiate cortical representations outside MLT during retrieval \cite{RuggVilberg13}.  
%

@@

To test this inverse relationship between theta and alpha we choose a recent hippocampal dependent multi-item association study to base our computational and empirical investigations on \cite{HornerBisbyBushEtAl15} (see Figure \ref{fig:deev}).  In this study the authors used a previously developed paradigm (see \textcite{HornerBurgess13,HornerBurgess14}) in which they manipulated the binding of a set of overlapping paired associates.   Here two types of associative stimuli were studied, closed-loop and open-loop. In the `closed-loop' events participants studied a set of paired associates on separate trials, e.g. Person-Location, Location-Object, and Object-Person, where each category refers to the same stimulus for a given event.  This created a complete set of binding between all elements.   In contrast, the `open-loop' events incorporated a new element and left some relationships unstudied, e.g. Object-Location, Location-Person, Person-Animal.  Each of the associations was later tested in a cued recognition task, which showed greater inter-event Dependency for the closed-loop compared to the open-loop associates.  Dependency here is operationalized as the retrieval of \emph{all} associated pairs from a given event, or none; whereas no Dependency would imply retrieval of some of the associated pairs but not all.  The goal of this metric is to capture the strength of the binding between elements within a given event, where high Dependency indicates strong associations leading to retrieval of all items within a given event, and low Dependency indicates weak associations leading to some of the items being retrieved incorrectly. In contrast, a non-associative measure of accuracy can be made using an Independent model which calculates Dependency based on the assumption that each retrieval trial is statistically independent from the next, see Section \ref{sec:CalcDep} for more details.  %This provides a nice measure of both Recollection yielding high Dependency from strong associations between elements, but also uses a recognition paradigm where successful retrieval with low between element Dependency implies a similar construct to Familiarity, i.e. retrieval without detailed recall.

@@

Building off our previous connectionist model of the hippocampus \cite{KetzMorkondaOReilly13}, referred to here as the Theta-Phase model, we seek to incorporate the ability to attend to particular dimensions of a given input pattern, while not attending to others.  Practically, this requires that each input be described on all possible stimulus dimensions, otherwise the hippocampal model will attempt to `recall' the missing information for those missing dimensions.  For example, one valid cortical representation experienced will contain information about size and shape but never in color, while another equally valid representation will contain information about color and shape, but never size.  The neural substrate supporting these representations needs to be able to represent all three dimension (and potentially many others) simultaneously and independently, but also be able to understand the inherent bias in which dimensions are likely to covary together.

%This problem begins to touch on the issue of intentional vs. incidental retrieval.  Intentional retrieval implies a goal directed, attentionally mediated re-activation of previously experienced representations, while incidental retrieval implies the reactivation of representations that are likely to be experienced together.  The Theta-Phase model is built exclusively on incidental retrieval mechanisms and currently has no way of reflecting the attentional, or goal directed, focus of the retrieval process.  Similarly, the lack of an intentional component to the retrieval process requires the Theta-Phase model to have valid, useful information in all potential stimulus dimensions, which is simply not possible in every scenario.  However, it is also crucial to maintain this incidental component to the model as it is fundamental to the Complimentary Learning Systems theory, and the main mechanism behind pattern completion \cite{McClellandMcNaughtonOReillyEtAl92,NormanOReilly03}.  What is required is an ability to balance between intentional and incidental retrieval drives in such a way that fits with known behavioral patterns.

A simple and ecologically valid way to reconcile this discrepancy is a mechanism that would highlight which of the stimulus dimensions are informative, or relevant given the current input.  This would be very similar to a selective attention mechanism.  Building from the evidence in support of an alpha mediated thalamic gating mechanism, we will explore modulation of cortical activity via thalamo-cortical interactions in the alpha band as a mechanism for the Theta-Phase model to gate specific stimulus dimensions for targeted encoding and retrieval.  The behavioral and physiological consequences of this mechanism will then be explored in a long-term memory paradigm that utilizes multi-element traces, and a bottom-up attentional signal to focus on specific stimulus dimensions.

@@

The other critical aspect of this bottom-up attention mechanism is the connection to alpha oscillations.  Historically, alpha power has been shown to correlate with the onset of sleep, which led to the suggestion that high alpha power indicates an idle or perhaps inhibited cortical region.  The mechanistic connection between alpha power and thalamic gating is captured in the firing properties of the thalamic relay cells (TRC) which show two distinct modes of firing.  The first shows oscillatory firing patterns acting like a pacemaker system in the alpha frequency range (approximately 10 Hz or 100ms periodicity), and is generally referred to as bursting.  This bursting mode is believed to be engage when a given cortical region is in an idle state, i.e. when no active processing is occurring in the thalamus or the corresponding cortex linked to that thalamic relay region.  The other regime, referred to as tonic firing, shows no specific oscillatory firing pattern, and instead shows increased spontaneous firing in both the thalamus and corresponding cortical regions \cite{LerescheLightowlerSolteszEtAl91,ShermanGuillery06,LopesdaSilva91}.  These dynamics have been studied extensively as a naturally occurring phenomena of thalamic interactions with cortex, however simulation studies have also shown the potential for these mechanisms to be used in a goal directed way \cite{HindriksvanPutten13,VijayanKopell12,LeeWhittingtonKopell13}. The implications of these results are that cortical regions being driving by TRCs in bust mode will show more alpha power, and will process the incoming cortical-cortical information less veridically, while cortical regions being driven by tonic TRC firing will show a relative decrease in alpha power and process cortical-cortical information more completely.  This is the foundation for the connection between alpha power and thalamic gating, such that increased alpha power is indicative of decreased cortical processing.

The detailed biology of this framework is well captured by \textcite{ShermanGuillery06}, and relies on two critical components not yet implemented in the Theta-Phase model: 1. The thalamic relay cells (TRC) that pass the gating signal onto the next cortical region, and 2. The cortical neurons in layers 5 and 6 that send this gating signal to the next TRC, and trigger the opening of the gate for processing in associated cortex for that TRC.  The following sections provides a high-level description of how these components are implemented, while supplementary info section \ref{appDeepLeabra} illustrates the detailed implementation into the LEABRA framework.

\begin{figure}
	\includegraphics[width=\textwidth]{figs/depNet.pdf}
        \caption[Thalamic Gating as Implemented in Theta-Phase Model] {  \emph{(A)} Schematic of Theta-Phase model with thalamic gating.  Posterior cortex connects with both layers 2/3 of Entorhinal cortex (EC) via cortical-cortical projects, but also to thalamic relay cells (TRC) via layer 5 bursting cells.  These TRC units maintain their topography into EC layer 4 where they provide a gating signal for EC to accept or suppress the corresponding cortical-cortical projection. Topographic connectivity shown in dashed arrows, and distributed connectivity shown in solid.  \emph{(B)}  Neural network showing TRC implementation as single units for a given cortical area's input and corresponding unit group in EC.  This topography is kept consistent between TRC and EC unit groups.  \emph{(C)} Conceptualization of EC unit groups to paired associate elements and thalamic gating signal manifested in EC with 2 bottom-up cortical inputs missing (i.e. object and animal) in unit groups with a low (red) gating signal, and 6 inputs present (i.e. location, person, and 4 contextual elements) shown with a high (yellow) gating signal.  The cortical input topography is maintained between EC and TRC where TRC activation is used to calculate the normalized gating signal shown in the corresponding EC unit groups.  Note the gating signal is not representation specific, but rather generalized for a given unit group. 
        }
     \label{fig:depNet}
\end{figure}

@@

In this work we start with the Theta-Phase model of the hippocampus and entorhinal cortex, as described in \textcite{KetzMorkondaOReilly13} and implemented in the Emergent framework \cite{AisaMingusOReilly08} (see Figure \ref{fig:depNet}).  Briefly, this model is built upon a series of structural and functional hypotheses based on anatomical and physiological data, which have been captured in the complementary learning systems (CLS) model of the hippocampus \cite{McClellandMcNaughtonOReilly95, NormanOReilly03}. The Entorhinal Cortex (EC) in the model is assumed to be the cortical gateway to the hippocampus.  This gateway feeds through the tri-synaptic pathway to the Dentate Gyrus (DG), CA3, and then to CA1.  Similarly, there is a parallel connection through the mono-synaptic pathway from the EC to the CA1 (and back).

The tri-synaptic pathway connections are broadly diffuse, and support the conjunctive binding of various distributed pieces of information into an overall episodic memory representation in the CA3. The mono-synaptic pathway conversely is topologically organized, not diffuse.  We capture this connectivity by organizing the simulated neurons in EC and CA1 into mutually interconnected {\em slots}, presumably encoding different separable elements/stimulus dimensions that loosely correspond to independent cortical regions converging on the EC \cite{Witter10}.  This slot architecture enables the mono-synaptic pathway to develop separable invertible pathways where a given EC input pattern can be encoded over a sparser representation in the corresponding CA1 slot, and this CA1 representation can in turn recover the full original EC slot pattern.

>>>>

% !TEX TS?program = pdflatexmk

%Dear all,
%We ar writing to invite you to submit an article to an upcoming special issue of the journal Hippocampus with the tentative title: Computational models of the hippocampus and related structures.  The theme includes computational and theoretical models as well as theory-driven data analysis of the hippocampus and related cortical and sub-cortical structures. 
%
%Articles for the special issue will be due June 15, 2019 with a planned publication date in early 2020.  All articles will be published on-line when accepted, but we will need to be firm about the deadline in order to avoid delays in publishing the full issue with volume and page numbers.
% 
%
%The contributions to the special issue can be in all forms of manuscripts accepted by Hippocampus (Commentaries, Reviews, Articles, Rapid Communications).
% 
%
%As is common with these types of special issues, we ask that any contributing author commit to reviewing two articles for the special issue.
% 
%
%Please let me know by Dec. 15 if you plan to contribute to this special issue.  Please also send me a tentative title and potential author list (which can change later if necessary).
%
%Best regards,
%Michael Hasselmo
%and Ehren Newman
% !BIB TS-program = biber

@@

\usepackage{subcaption}

\usepackage[american]{babel}
\usepackage{csquotes}
\usepackage[style=apa]{biblatex}
\DeclareLanguageMapping{american}{american-apa}

\addbibresource{references.bib}
\usepackage{longtable}
\usepackage{csvsimple}				% for reading cvs files into tables

@@

\section*{Acknowledgements}\label{sec:acknowledge}
The authors would like to thank research assistants at Boulder, especially Chris Bird, Krystin Corby, Shaina Martis, as well as other undergraduate volunteers for help in collecting the data used in the EEG experiments.  We would also like to thank faculty at Boulder, specifically McKell Carter, Anu Sharma and Erik Willcutt for helpful feedback during manuscript preparation.  The authors have no known conflicts of interest with this work.  This work was partially funded by NSF Research Fellowship grant \#DGE-0707432.

@@

It is generally accepted that long-term memory encoding and retrieval are not isolated functions of any one brain area \parencite{Gabrieli98,PrinceDaselaarCabeza05}.  However, there are many challenges in moving beyond the traditional studies focusing on individual regions  (e.g., concentrating on the prominent role of the hippocampus), to obtain insight into the nature of network-level dynamics.  An increasing body of work support the idea that oscillatory dynamics in multiple brain regions are associated with different aspects of  memory encoding and retrieval \parencite{NyhusCurran10,HanslmayrStaudiglFellner12,Klimesch12}, %update refs here!!!
These studies have generally found a consistent relationship between several frequency bands and successful encoding or retrieval of experiences; namely that oscillatory power within the alpha (8 to 12Hz) and beta (13 to 30Hz) frequency bands generally decrease with the successful execution of these memory processes, while theta (3 to 8Hz) and gamma (30 to 100+Hz) increase in the regions directly involved in the memory operations. Why these relationships exist, and what they tell us about the larger functional organization of memory systems in the brain are still open questions.

One model suggests that thalamic circuits play a functional role in selectively gating communication between cortical regions by synchronizing them in various frequency bands \parencite{ShermanGuillery06,SaalmannKastner11,KetzJensenOReilly15}.  These gating mechanisms, implemented  by oscillatory synchronization, are central to system level brain function and ultimately affect cognitive behavior such as memory encoding and retrieval.  Further, the particular oscillatory dynamics engaged by this selection process may provide indicators of which system-scale neural circuits are coordinating together to facilitate that particular form of memory processing.

A particular thalmo-cortical circuit of interest for memory processes involves the hippocampus and its associated subcortical network, including the anterior thalamus, and medial septum which are essential in the maintenance of theta oscillations (4 to 8 Hz) \parencite{Buzsaki02}.  Theta oscillations have been shown to be critically involved in successful encoding and retrieval of episodic like memories in both humans and animals \parencite{NyhusCurran10,JonesWilson05,FuentemillaBarnesDuzelEtAl14}, and are believed to be primarily driven by the pacemaker cells within the medial septum.  The anterior thalamus, however,  provides the connectivity for the hippocampus to potentially modulate communication with the medial and orbital PFC through theta synchronization \parencite{VanderWerfJollesWitterEtAl03,AggletonDumontWarburton11}.  Thus, the theta circuit is important for the core functioning of the hippocampus during encoding and retrieval.

%There is MEG evidence that theta synchronization with the hippocampus can facilitate Recollective memory processes in humans \parencite{FuentemillaBarnesDuzelEtAl14}. There is some indirect MEG and BOLD evidence that the thalamus plays a role in this synchronization process \parencite{FuentemillaBarnesDuzelEtAl14,PergolaRanftMathiasEtAl13}, which is bolstered by the known connectivity between the hippocampus and PFC through the anterior thalamus from rodents \parencite{AggletonBrown99,AggletonDumontWarburton11}.  The thalamic gating model connects these ideas suggesting Recollective memory works through the hippocampus via theta synchronization, however to date there is no strong evidence supporting this claim in humans.

A complimentary thalamic circuit involves the pulvinar providing synchronization within the alpha (8 to 12 Hz) band between parietal and visual cortices with the medial, orbital and lateral PFC \parencite{BarbasHenionDermon91,KievitKuypers77,LopesdaSilva91}.  This circuit is important for bottom-up stimulus-driven encoding processes in memory \parencite{JutrasFriesBuffalo13}, but it can also inhibit contradictory or interfering information during internally guided retrieval \parencite{ParkLeeKangEtAl14}.  Thus, its overall contribution to memory can be mixed, and depends on the nature of the task and what kinds of memory processing is required.

%The pulvinar has a well established role in coordinating alpha synchrony in animal models \parencite{SaalmannKastner11,SaalmannKastner09,LopesdaSilva91,LopesdaSilvaVosMooibroekEtAl80}, and there is recent evidence in human models showing alpha oscillations provide an active function of suppressing cortex that would otherwise interfere with the memory retrieval \parencite{WaldhauserJohanssonHanslmayr12,HanslmayrStaudiglAslanEtAl10,KhaderRosler11}.  There is also human fMRI evidence showing PFC modulation of visual cortex through the pulvinar \parencite{SotoHumphreysRotshtein07,RotshteinSotoGrecucciEtAl11}.

Indeed, a recent spiking model of these oscillations called the Sync/deSync Model highlights the role of a synchronized hippocampus that accompanies a desynchronized neocortex during encoding and retrieval.  This model in particular focuses on the role of decreased alpha power related to successful encoding of new information \parencite{ParishHanslmayrBowman18, HanslmayrStaresinaBowman16}.  Similarly, recent empirical studies of human intracranial recordings find evidence to support the complimentary roles of hippocampal synchrony and neocortical desynchrony in successful encoding and retrieval \parencite{GriffithsMichelmannRouxEtAl18, StaresinaMichelmannBonnefondEtAl16}.  In general these results find a decrease in neocortical alpha power and an increase in hippocampal gamma power that coincide with successful encoding and retrieval of declarative memories.

This accumulated evidence suggests a relationship between theta and alpha frequency bands and long-term memory systems as constructed by the Complimentary Learning Systems theory \parencite{McClellandMcNaughtonOReilly95,OReillyBhattacharyyaHowardEtAl14}.  Namely, that theta power positively tracks successful engagement (encoding/retrieval) between hippocampus and neocortex, while alpha power negatively tracks with this process as active inhibition of cortex is released in response to the exchange of information between hippocampus and neocortex.  This relationship is further supported by the connectivity of the thalamus with cortex such that a specific model for how information is routed into and out of these systems can be tied to empirical measures of neural oscillations.  This work seeks to test this relationship through a series of computational simulations of hippocampal-thalamo-cortical interaction, and an accompanying EEG study in humans to compare against simulation results.

@@

\subsection{Dependent Events Paradigm}
%
%Hippocampal dependent memory is generally thought to be associative in nature, where multiple cortical representations are bound together in a distinct hippocampal representation during encoding.  Later, an incomplete set of those cortical representations can be used as a cue to reactive the full set during retrieval \parencite{McClellandMcNaughtonOReillyEtAl92}.  This multi-element type of memory retrieval is also referred to behaviorally as `Recollection' where multiple details of a given experience can be retrieved, and hippocampal activity increases with subjective confidence and the amount of information being retrieved \parencite{RuggVilbergMattsonEtAl12}.  
%
%In contrast to Recollection, memories that exhibit `Familiarity' have no measurable details that can be recalled, however individual elements can be successfully recognized when presented with a mix of lures.  This Familiarity process is thought to rely more heavily on the medial temporal lobe (MTL) cortex surrounding the hippocampus, including entorhrinal, parahippocampal, and perirhinal cortices.  These cortical areas, among others, are proposed to be a separate memory system from the hippocampus, serving behavior when recall of all associated details is not necessary or available \parencite{EichenbaumYonelinasRanganath07,AggletonBrown99}.  This theory of separate memory systems has been contentious for several decades with evidence both in support and in contrast to its core ideas  \parencite{EichenbaumYonelinasRanganath07,WixtedSquire10}; namely that Recollection can occur in addition to Familiarity and that the hippocampus is not specific to Recollection but can be used in Familiarity based processing.  What is clear is that MTL cortex learns and maintains representations required for hippocampal retrieval, and those MTL regions along with the hippocampus can be used to re-instantiate cortical representations outside MLT during retrieval \parencite{RuggVilberg13}.  
%

@@

To test this inverse relationship between theta and alpha we choose a recent hippocampal dependent multi-item association study to base our computational and empirical investigations on \parencite{HornerBisbyBushEtAl15} (see Figure \ref{fig:deev}).  In this study the authors used a previously developed paradigm (see \textcite{HornerBurgess13,HornerBurgess14}) in which they manipulated the binding of a set of overlapping paired associates.   Here two types of associative stimuli were studied, closed-loop and open-loop. In the `closed-loop' events participants studied a set of paired associates on separate trials, e.g. Person-Location, Location-Object, and Object-Person, where each category refers to the same stimulus for a given event.  This created a complete set of binding between all elements.   In contrast, the `open-loop' events incorporated a new element and left some relationships unstudied, e.g. Object-Location, Location-Person, Person-Animal.  Each of the associations was later tested in a cued recognition task, which showed greater inter-event Dependency for the closed-loop compared to the open-loop associates.  Dependency here is operationalized as the retrieval of \emph{all} associated pairs from a given event, or none; whereas no Dependency would imply retrieval of some of the associated pairs but not all.  The goal of this metric is to capture the strength of the binding between elements within a given event, where high Dependency indicates strong associations leading to retrieval of all items within a given event, and low Dependency indicates weak associations leading to some of the items being retrieved incorrectly. In contrast, a non-associative measure of accuracy can be made using an Independent model which calculates Dependency based on the assumption that each retrieval trial is statistically independent from the next, see Section \ref{sec:CalcDep} for more details.  %This provides a nice measure of both Recollection yielding high Dependency from strong associations between elements, but also uses a recognition paradigm where successful retrieval with low between element Dependency implies a similar construct to Familiarity, i.e. retrieval without detailed recall.

@@

Building off our previous connectionist model of the hippocampus \parencite{KetzMorkondaOReilly13}, referred to here as the Theta-Phase model, we seek to incorporate the ability to attend to particular dimensions of a given input pattern, while not attending to others.  Practically, this requires that each input be described on all possible stimulus dimensions, otherwise the hippocampal model will attempt to `recall' the missing information for those missing dimensions.  For example, one valid cortical representation experienced will contain information about size and shape but never in color, while another equally valid representation will contain information about color and shape, but never size.  The neural substrate supporting these representations needs to be able to represent all three dimension (and potentially many others) simultaneously and independently, but also be able to understand the inherent bias in which dimensions are likely to covary together.

%This problem begins to touch on the issue of intentional vs. incidental retrieval.  Intentional retrieval implies a goal directed, attentionally mediated re-activation of previously experienced representations, while incidental retrieval implies the reactivation of representations that are likely to be experienced together.  The Theta-Phase model is built exclusively on incidental retrieval mechanisms and currently has no way of reflecting the attentional, or goal directed, focus of the retrieval process.  Similarly, the lack of an intentional component to the retrieval process requires the Theta-Phase model to have valid, useful information in all potential stimulus dimensions, which is simply not possible in every scenario.  However, it is also crucial to maintain this incidental component to the model as it is fundamental to the Complimentary Learning Systems theory, and the main mechanism behind pattern completion \parencite{McClellandMcNaughtonOReillyEtAl92,NormanOReilly03}.  What is required is an ability to balance between intentional and incidental retrieval drives in such a way that fits with known behavioral patterns.

A simple and ecologically valid way to reconcile this discrepancy is a mechanism that would highlight which of the stimulus dimensions are informative, or relevant given the current input.  This would be very similar to a selective attention mechanism.  Building from the evidence in support of an alpha mediated thalamic gating mechanism, we will explore modulation of cortical activity via thalamo-cortical interactions in the alpha band as a mechanism for the Theta-Phase model to gate or modulate specific stimulus dimensions for targeted encoding and retrieval.  The behavioral and physiological consequences of this mechanism will then be explored in a long-term memory paradigm that utilizes multi-element traces, and a bottom-up attentional signal to focus on specific stimulus dimensions.

@@

The other critical aspect of this bottom-up attention mechanism is the connection to alpha oscillations.  Historically, alpha power has been shown to correlate with the onset of sleep, which led to the suggestion that high alpha power indicates an idle or perhaps inhibited cortical region.  The mechanistic connection between alpha power and thalamic gating is captured in the firing properties of the thalamic relay cells (TRC) which show two distinct modes of firing.  The first shows oscillatory firing patterns acting like a pacemaker system in the alpha frequency range (approximately 10 Hz or 100ms periodicity), and is generally referred to as bursting.  This bursting mode is believed to be engage when a given cortical region is in an idle state, i.e. when no active processing is occurring in the thalamus or the corresponding cortex linked to that thalamic relay region.  The other regime, referred to as tonic firing, shows no specific oscillatory firing pattern, and instead shows increased spontaneous firing in both the thalamus and corresponding cortical regions \parencite{LerescheLightowlerSolteszEtAl91,ShermanGuillery06,LopesdaSilva91}.  These dynamics have been studied extensively as a naturally occurring phenomena of thalamic interactions with cortex, however simulation studies have also shown the potential for these mechanisms to be used in a goal directed way \parencite{HindriksvanPutten13,VijayanKopell12,LeeWhittingtonKopell13}. The implications of these results are that cortical regions being driving by TRCs in bust mode will show more alpha power, and will process the incoming cortical-cortical information less veridically, while cortical regions being driven by tonic TRC firing will show a relative decrease in alpha power and process cortical-cortical information more completely.  This is the foundation for the connection between alpha power and thalamic gating, such that increased alpha power is indicative of decreased cortical processing.

The detailed biology of this framework is well captured by \textcite{ShermanGuillery06}, and relies on two critical components not yet implemented in the Theta-Phase model: 1. The thalamic relay cells (TRC) that pass the gating signal onto the next cortical region, and 2. The cortical neurons in layers 5 and 6 that send this gating signal to the next TRC, and trigger the release from inhibition for processing in associated cortex for that TRC.  The following sections provides a high-level description of how these components are implemented, while supplementary info section \ref{appDeepLeabra} illustrates the detailed implementation into the LEABRA framework.

\begin{figure}
	\includegraphics[width=\textwidth]{figs/depNet_edit2.pdf}
        \caption[Thalamic Gating as Implemented in Theta-Phase Model] {  \emph{(A)} Schematic of Theta-Phase model with thalamic gating.  Posterior cortex connects with both layers 2/3 of Entorhinal cortex (EC) via cortical-cortical projects, but also to thalamic relay cells (TRC) via layer 5 bursting cells.  These TRC units maintain their topography into EC layer 4 where they provide a gating signal for EC to accept or suppress the corresponding cortical-cortical projection. Topographic connectivity shown in dashed arrows, and distributed connectivity shown in solid.  \emph{(B)}  Neural network showing TRC implementation as single units for a given cortical area's input and corresponding unit group in EC.  This topography is kept consistent between TRC and EC unit groups.  \emph{(C)} Illustration of how TRC modulates  network activity during cued-recognition.  Here, a particular person (labeled Per) pattern is being used to cue recognition of an associated animal (labeled Ani) pattern, as indicated by the labels on posterior cortex unit groups.  This input pattern from posterior cortex influences ECin and TRC cells releasing the thalamic suppression in unit groups where there is sufficient activation (i.e. the person and animal unit groups).  The activation from ECin flows up to the hippocampus where missing information can be retrieved, and comes back to ECout through connections from CA1.  ECout's activation influences both ECout and the TRC activations, using the hippocampally retrieved information to fill in the animal unit group as well as show partial recall of a non-target association of location pattern.  The cortical input topography is maintained between EC and TRC where TRC activation is used to calculate the normalized gating signal shown in the corresponding EC units.  Note the gating signal is not representation specific, but rather multiplied across each of the units of its target unit group. 
        }
     \label{fig:depNet}
\end{figure}

@@

In this work we start with the Theta-Phase model of the hippocampus and entorhinal cortex, as described in \textcite{KetzMorkondaOReilly13} and implemented in the Emergent framework \parencite{AisaMingusOReilly08} (see Figure \ref{fig:depNet}).  Briefly, this model is built upon a series of structural and functional hypotheses based on anatomical and physiological data, which have been captured in the complementary learning systems (CLS) model of the hippocampus \parencite{McClellandMcNaughtonOReilly95, NormanOReilly03}. The Entorhinal Cortex (EC) in the model is assumed to be the cortical gateway to the hippocampus.  This gateway feeds through the tri-synaptic pathway to the Dentate Gyrus (DG), CA3, and then to CA1.  Similarly, there is a parallel connection through the mono-synaptic pathway from the EC to the CA1 (and back).

The tri-synaptic pathway connections are broadly diffuse, and support the conjunctive binding of various distributed pieces of information into an overall episodic memory representation in the CA3. The mono-synaptic pathway conversely is topologically organized, not diffuse.  We capture this connectivity by organizing the simulated neurons in EC and CA1 into mutually interconnected {\em slots}, presumably encoding different separable elements/stimulus dimensions that loosely correspond to independent cortical regions converging on the EC \parencite{Witter10}.  This slot architecture enables the mono-synaptic pathway to develop separable invertible pathways where a given EC input pattern can be encoded over a sparser representation in the corresponding CA1 slot, and this CA1 representation can in turn recover the full original EC slot pattern.

The learning dynamics within this model, as governed by the LEABRA algorithm, are dictated by the phases of a presumed theta oscillation, thus its short-hand name of 'Theta-Phase' used here. In short, during learning two error signals are derived within every period of a give theta oscillation.  The first, during the trough of the theta oscillation, provides an error driven learning signal to the mono-synaptic pathway, learning the invertible mapping between EC and CA1 can be thought of as an encoding.  The second, during the peak of the theta oscillation, provides a error driven learning signal on the tri-synaptic pathway, learning to recollect missing or incomplete information can be thought of as retrieval.  To be clear, both pathways are providing an active learning signal to 'encode' information.  Similarly, while the tri-synaptic pathway contributes to retrieval of associations between unit groups, the mono-synaptic pathway can contribute to retrieval within unit groups. The overall contributions of either pathway to learning in general has been illustrated in a recent study \parencite{SchapiroEtAl17}.
